---
title: "!Rmd_02_forecasting.Rmd"
author: "Diego Uchendu Learning from Prof Dr Thomas Kirschstein"
date: "01/12/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Time series demand example

```{r}
library(tidyverse)
library(forecast)
library(expsmooth)
library(TTR)
library(fpp)
library(fpp2)

```

**Constant**

```{r}
len <- 100

y.const <- rep(50, len) # replicates value 50 100 times



x <- 1:len #1 to 100

#png(file = "forecast_ts_const.png", bg = "transparent", width=600, height = 400)
#win.metafile("forecast_ts_const.wmf", width=9, height = 6)
par(family="serif", mar = c(5,5,.1,.1))
plot(x , y.const , type = "l", xlab= "time/periods", ylab ="demand", lwd = 1.5 , cex.lab = 2, cex.axis = 1.5, main= "Constant")
#dev.off()

```

**TREND**

```{r}
# TREND

y.trend <- 1:len*0.1 #between 0.1 to 10

#png(file = "forecast_ts_trend.png", bg = "transparent", width=600, height = 400)
#win.metafile("forecast_ts_trend.wmf", width=9, height = 6)
par(family="serif", mar = c(5,5,.1,.1))
plot(x , y.trend , type = "l", xlab= "time/periods", ylab ="demand", lwd = 1.5 , cex.lab = 2, cex.axis = 1.5,main="Trend")
#dev.off()
```


**Seasonality**


```{r}
#SEASONALITY

x.seas <- 0:(len-1)%%4 + 1 # creates seasonal values. 1,2,3,4 multiple times.

#drops as low as 0 and peaks as high as 6, {0,2,4,6} repeated.
y.seas <- (x.seas-1)*2 # quarterly fluctuations

#png(file = "forecast_ts_seas.png", bg = "transparent", width=600, height = 400)
# win.metafile("forecast_ts_seas.wmf", width=9, height = 6)
par(family="serif", mar = c(5,5,.1,.1))
plot(x , y.seas , type = "l", xlab= "time/periods", ylab ="demand", lwd = 1.5 , cex.lab = 2, cex.axis = 1.5, main= "Seasonality")
#dev.off()

```


**Error**

```{r}
#ERROR

y.err <- rnorm(len, 0 , 5) 

# win.metafile("forecast_ts_err.wmf", width=9, height = 6)
par(family="serif", mar = c(5,5,.1,.1))
plot(x , y.err , type = "l", xlab= "time/periods", ylab ="demand", lwd = 1.5 , cex.lab = 2, cex.axis = 1.5, main="Error")

#dev.off()
```

**Constant + Trend + Seasonality**

```{r}

#png(file = "forecast_ts_all.png", bg = "transparent", width=600, height = 400)
# win.metafile("forecast_ts_all.wmf", width=9, height = 6)
par(family="serif", mar = c(5,5,.1,.1))
plot(x , y.const + y.trend + y.seas , type = "l", xlab= "time/periods", ylab ="demand", lwd = 1.5 , cex.lab = 2, cex.axis = 1.5, main = "Constant + Trend + Seasonality")
#dev.off()
```


**Error + Constant + Trend + Season**

```{r}
# win.metafile("forecast_ts_all_err.wmf", width=9, height = 6)
par(family="serif", mar = c(5,5,.1,.1))
plot(x ,y.err +  y.const + y.trend + y.seas , type = "l", xlab= "time/periods", ylab ="demand", lwd = 1.5 , cex.lab = 2, cex.axis = 1.5,main="Error + Constant + Trend + Season")
#dev.off()
```

The better the forecast, the smaller is the forecasting error $\epsilon_{it}$ for material $i$ in a particular time period $t$. The forecast error determines the safety stock level. $\epsilon_{it}$ follows a stochastic process.

**Time series values**

```{r}
# ts values
y.com <- y.err +  y.const + y.trend + y.seas
#y.com # 100 values

# deltat is the fraction of the sampling period between successive observations; e.g., 1/12 for monthly data.
# 1/4 for quarterly data
y.com.ts <- ts(y.com, deltat = 1/4) # deltat divides the 100 values into 4 parts.

y.com.ts
```

**TS decomosition by stl**

seasonal, trend and irregular components using loess.

```{r}
stl.y <- stl(y.com.ts, s.window = 4)
plot(stl.y)
```


**TS decomposition by decompose**

decomposition of additive time series.

```{r}
decompose.y <- decompose(y.com.ts)
plot(decompose.y)
```

**TS decomosition by lm**

lm is linear models, It can be used to carry out regression, single stratum analysis of variance and analysis of covariance.

x.seas <- 0:(len-1)%%4 + 1 # creates seasonal values. 1,2,3,4 multiple times.

y.seas <- (x.seas-1)*2 # quarterly fluctuations

**Regression Approach**

* $y_t:$ demand in period $t=\{1,...,T\}$.
* $o \in \mathbb N$ the periodicity, the season of a period $t$ is defined as $s_t \in \{1,...,o\}$.
* $x^t=(1,...,T)$ denotes the time vector.
* $x^s=(s_1,...,s_T)$ the seasonality vector.

$$ y_t= \beta_0 + \beta_{st} + \beta_1 \cdot t + \epsilon_t$$

* $o$ represents seasonality.
* if $o=12$ means seasons in months.
* $o=4$ seasons  Quarterly.
* $o=30$ seasons for days in a month.
* $o=7$ seasons for days in a week.

All $\beta$ parameters are to be estimated.

* $\beta_{st}$ additive seasonal effect.
* $\beta_1$ slope parameter representing linear trend increasing or decreasing over time.

We usually run test whether one of these $\beta_0, \beta_{st},\beta_1$ is significantly different from 0.

* We can use T-test or do complete regression analysis, so the complete model is explaining something.

* f-test look at the individual model and try to reduce that.

* We can reduce that using stepwise regression technique, backward regression starting with the full model and reducing the parameter iteratively by backwards algorithm.


```{r}
# TS decomosition by lm
#x.seas is seasonal ordering data 
#x is the main data
#y.com is time series values y.err +  y.const + y.trend + y.seas

# as.factor meaning x.seas is ordered, as.factor(x.seas) indicates 4 levels
as.factor(x.seas)
```


```{r}
y.lm.decom <- lm(y.com ~ x + as.factor(x.seas) )
y.lm.decom
```



```{r}
summary(y.lm.decom)
```




```{r}
anova(y.lm.decom)
```


Predicted and residual

```{r}
y.lm.pred <- predict(y.lm.decom)
y.lm.res <- residuals(y.lm.decom)
```

# Errors

**MSE**
Mean Squared Error

$\epsilon_t^2=(y_t-\hat y_t)^2$

$$MSE=\frac {1}{T} \sum_{t=1}^T \epsilon^2_t$$


```{r}
# MSE
sum(y.lm.res^2)/100
```

**MAD**

Median absolute deviation (MAD).

$|\epsilon_t|=|y_t-\hat y_t|$

$$ meadian(|\epsilon|)$$
```{r}
median(abs(y.lm.res))
```

**Absolute percentage error**

$\frac{|\epsilon_t|}{y_t}=\frac {|y_t -\hat y_t|}{y_t}$

$$MAPE= \frac {1}{T} \sum_{t=1}^T \frac {|\epsilon_t|}{y_t} $$

```{r}
# MAPE
mean(abs(y.lm.res)/y.lm.pred)
```


**Mean squared log accuracy ratio**

$q_t=In(\frac{\hat y_t}{y_t})^2$

$$\frac {1}{T} \sum_{t=1}^T q_t$$


```{r}
# log q
mean(log(y.lm.pred/y.com)^2)
```


**function for calculating accuracy measures**

```{r}
fc.stats <- function(yhat, y){
  y.res <- y -  yhat
  # MSE
  mse <- mean(y.res^2, na.rm =T)
  # MAD
  mad <- median(abs(y.res), na.rm =T)
  # MAPE
  mape <- mean(abs(y.res)/yhat, na.rm =T)
  # log q
  msla <- mean(log(yhat/y)^2, na.rm =T)
  #   return
  res <- c(mse,mad,mape,msla)
  names(res) <- c("MSE","MAD","MAPE","MSLA")
  return(res)
  }

# test functions
fc.stats(yhat = y.lm.pred, y = y.com)
```


# Regression analysis of examplary time series  ##############################


* $\beta_1$ is trend estimate parameter
* $y_{t,trend}$ is the true demand value containing only the trend which is the $y.trend$
* $\hat y_t$ estimate demand value which is  $y.lm.trend$
* $x$ is the time/ period from 1 to 100.

$$\hat y_t= \beta_1* \vec x$$


```{r}
#
y.lm.trend <- coefficients(y.lm.decom)[2] * x


#win.metafile("forecast_ts_lm_trend.wmf", width=9, height = 6)
par(family="serif", mar = c(5,5,.1,.1))
plot(x, y.lm.trend, type="l", xlab= "time/periods", ylab ="demand", lwd = 1.5 , cex.lab = 2, cex.axis = 1.5)
lines(x, y.trend, type="l", col="red", lwd = 1.5)
legend("topleft", col=c("red","black"), legend = c("true model","estimate"), lwd=c(2,2), cex=1.75, bg ="white")
#dev.off()
```

$\beta_{02 is }$ ```{r} coefficients(y.lm.decom)[3]```

$ y_{t,seas}= \beta_{01}, \beta_{02}, \beta_{03}, \beta_{04}$ replicated lenght of $\frac {lenght|\vec x|}{4}$ from 1 to lenght $\vec x$

**Please Estimate beta01 correctly**


```{r}
# tail(coefficients(y.lm.decom),3) means extract the last 3 values from the y.lm.decom vector

c(0,tail(coefficients(y.lm.decom),3))
```

```{r}
lm.seas.coeff <- c(0,tail(coefficients(y.lm.decom),3))
y.lm.seas <- rep(lm.seas.coeff, ceiling(len/4))[1:len]
#y.lm.seas
head(y.lm.seas)

```

```{r}
#y.lm.seas <- rep(lm.seas.coeff, ceiling(len/4))[1:len]

#win.metafile("forecast_ts_lm_seas.wmf", width=9, height = 6)
par(family="serif", mar = c(5,5,.1,.1))
plot(x, y.lm.seas, type="l", xlab= "time/periods", ylab ="demand", lwd = 1.5 , cex.lab = 2, cex.axis = 1.5)
lines(x, y.seas, type="l", col="red")
legend("topleft", col=c("red","black"), legend = c("true model","estimate"), lwd=c(2,2), cex=1.75, bg ="white")
#dev.off()
```

**estimated(predicted) vs actual**

y.com= y.err +  y.const + y.trend + y.seas


```{r}
#win.metafile("forecast_ts_lm_pred.wmf", width=9, height = 6)
par(family="serif", mar = c(5,5,.1,.1))
plot(x, y.com, type="l", xlab= "time/periods", ylab ="demand", lwd = 1.5 , cex.lab = 2, cex.axis = 1.5)
lines(x, y.lm.pred, type="l", col="red")
legend("topleft", col=c("red","black"), legend = c("estimate","observation"), lwd=c(2,2), cex=1.75, bg ="white")
#dev.off()
```


**Error**

y.error = actual error, thus actual variance.
residuals(y.lm.decom)= estimated error.

```{r}
#win.metafile("forecast_ts_lm_err.wmf", width=9, height = 6)
par(family="serif", mar = c(5,5,.1,.1))
plot(x, residuals(y.lm.decom), type="h", xlab= "time/periods", ylab ="residual", lwd = 1.25 , cex.lab = 2, cex.axis = 1.5, pch =15)
points(x-.25, y.err, type="h", col="red", lwd=1.25)
legend("topleft", col=c("red","black"), legend = c("estimate","observation"), lwd=c(2,2), cex=1.75, bg ="white")
#dev.off()
```



# moving average ############################## 
mn